---
title: "Spatial data visualization"
author: ""
date: ""
output: html_document
editor_options: 
  chunk_output_type: console
---

## Main ideas

- Spatial data is important
  - exploratory data analysis
  - detecting spatial patterns and trends
  - understanding spatial data relationships
  - analysis of spatial data should reflect spatial structure

## Keyboard shortcuts

| **Task / function**  | **Windows & Linux** | **macOS**           |
|----------------------|---------------------|---------------------|
| Insert R chunk       | Ctrl+Alt+I          | Command+Option+I    |
| Knit document        | Ctrl+Shift+K        | Command+Shift+K     |
| Run current line     | Ctrl+Enter          | Command+Enter       |
| Run current chunk    | Ctrl+Shift+Enter    | Command+Shift+Enter |
| Run all chunks above | Ctrl+Alt+P          | Command+Option+P    |
| `<-`                 | Alt+-               | Option+-            |
| `%>%`                | Ctrl+Shift+M        | Command+Shift+M     |

## Packages

```{r load-packages, message = FALSE}
library(tidyverse)
library(sf)
```

## Lecture notes and exercises

**Spatial data is different**

Our typical "tidy" dataframe.

```{r tidy-data}
mpg
```

A simple feature object.

```{r sf-data}
nc <- st_read("data/nc_covid.shp", quiet = TRUE)
nc
```

**Question:** What differences do you observe when comparing a typical tidy
data frame to the new simple feature object?

### Simple features

A **simple feature** is a standard, formal way to describe how real-world 
spatial objects (country, building, tree, road, etc) can be represented by a 
computer. 

The `sf` package implements simple features and other spatial functionality 
using **tidy** principles. Simple features have a geometry type. Common choices
are shown in the slides associated with today's lecture. 

Simple features are stored in a data frame, with the geographic information 
in a column called `geometry`. Simple features can contain both spatial and
non-spatial data.

The main functions in the `sf` package begin with `st_`.

### `sf` and `geom_sf()`

Notice `nc` contains both spatial and non-spatial information.

We can build up a visualization layer-by-layer beginning with `ggplot`. Let's 
start by making a basic plot of North Carolina counties.

```{r plot-1, fig.width=10, fig.height=4}
ggplot(nc) +
  geom_sf() +
  labs(title = "North Carolina counties")
```

Now adjust the theme with `theme_bw()`.

```{r plot-2, fig.width=10, fig.height=4}
ggplot(nc) +
  geom_sf() +
  labs(title = "North Carolina counties with theme") + 
  theme_bw()
```

Now adjust `color` in `geom_sf` to change the color of the county borders.

```{r plot-3, fig.width=10, fig.height=4}
ggplot(nc) +
  geom_sf(color = "darkgreen") +
  labs(title = "North Carolina counties with theme and aesthetics") + 
  theme_bw() 
```

Then increase the width of the county borders using `size`.

```{r plot-4, fig.width=10, fig.height=4}
ggplot(nc) +
  geom_sf(color = "darkgreen", size = 1.5) +
  labs(title = "North Carolina counties with theme and aesthetics") +
  theme_bw()
```

Fill the counties by specifying a `fill` argument.

```{r plot-5, fig.width=10, fig.height=4}
ggplot(nc) +
  geom_sf(color = "darkgreen", size = 1.5, fill = "orange") +
  labs(title = "North Carolina counties with theme and aesthetics") +
  theme_bw()
```

Finally, adjust the transparency using `alpha`.

```{r plot-6, fig.width=10, fig.height=4}
ggplot(nc) +
  geom_sf(color = "darkgreen", size = 1.5, fill = "orange", alpha = 0.50) +
  labs(title = "North Carolina counties with theme and aesthetics") +
  theme_bw()
```

Our current map is a bit much. Adjust `color`, `size`, `fill`, and `alpha` until
you have a map that effectively displays the counties of North Carolina.

### North Carolina COVID-19 Mapping

The `nc` data was pulled from the [New York Times COVID-19 Dashboard](https://www.nytimes.com/interactive/2020/us/north-carolina-coronavirus-cases.html) 
as of 02-04-2021.

The dataset contains the following variables on all North Carolina counties:

- `county`: county name
- `cases`: total number of COVID-19 cases
- `deaths`: total number of COVID-19 deaths
- `case100`: number of COVID-19 cases per 100,000
- `death100`: number of COVID-19 deaths per 100,000

Let's use the COVID-19 data to generate a **choropleth map**.

```{r covid-plot-1, fig.width=10, fig.height=4}
ggplot(nc) +
  geom_sf(aes(fill = cases)) + 
  labs(title = "Higher population counties have more COVID-19 cases",
       fill = "# cases") + 
  theme_bw() 
```

It is best to choose your own color palette. A great resource is
[colorbrewer2](http://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3).

One way to set fill colors is with `scale_fill_gradient()`.

```{r covid-plot-2, fig.width=10, fig.height=4}
ggplot(nc) +
  geom_sf(aes(fill = cases)) +
  scale_fill_gradient(low = "#fee8c8", high = "#7f0000") +
  labs(title = "Higher population counties have more COVID-19 cases",
       fill = "# cases") + 
  theme_bw() 
```

**Question:** Is the above visualization informative? Why or why not?

```{r fig.width=10, fig.height=4}
ggplot(nc) +
  geom_sf(aes(fill = case100)) +
  scale_fill_gradient(low = "#fff7f3", high = "#49006a") +
  labs(title = "COVID-19 cases per 100,000") +
  theme_bw()
```

### Challenges

1. Different types of data exist (raster and vector).

2. The coordinate reference system (CRS) matters.

3. Manipulating spatial data objects is similar, but not identical to
   manipulating data frames. 

### `dplyr`

The `sf` package plays nicely with our earlier data wrangling functions from 
`dplyr`.

## `select()`

```{r sf-and-select}
nc %>% 
  select(deaths, death100)
```

## `filter()`

```{r sf-and-filter}
nc %>% 
  filter(deaths > 100)
```

## `summarize()`

```{r sf-and-summarize}
nc %>% 
  summarize(total_deaths  = sum(deaths),
            total_cases   = sum(cases))
```

**Question:** Does the above geometry have any meaning?

Geometries are "sticky". They are kept until deliberately dropped using 
`st_drop_geometry`.

```{r drop-geometry}
nc %>% 
  select(county, deaths) %>% 
  filter(deaths > 100) %>% 
  st_drop_geometry()
```

## Practice

(1) Construct an effective visualization investigating the spatial distribution
of COVID-19 deaths in North Carolina. Carefully consider aesthetics and choose 
your own color palette using colorbrewer2.

```{r practice-1}
nc %>% 
  ggplot() +
  geom_sf(mapping = aes(fill = death100)) +
  scale_fill_gradient(low = "#fff7bc", high = "#d95f0e") +
  annotate(geom = "point", y = 36.0014, x = -78.9382, color = "purple", size = 3) +
  annotate(geom = "text", y = 36.15, x = -78.9382, label = "Duke", 
           color = "purple", size = 5) +
  theme_void() +
  theme(legend.position = "bottom") +
  labs(fill = "Deaths per\n100,000")
```

(2) Which counties have a relatively large number of deaths given their case 
count? Which counties have a relatively small number of deaths given their 
case count? Construct an effective visualization to answer this question and 
carefully consider all aesthetic choices.

```{r practice-2}
nc %>% 
  mutate(ratio = deaths / cases) %>% 
  ggplot() +
  geom_sf(mapping = aes(fill = ratio)) +
  scale_fill_gradient(low = "#fff7bc", high = "#d95f0e") +
  annotate(geom = "point", y = 36.0014, x = -78.9382, color = "purple", size = 3) +
  annotate(geom = "text", y = 36.15, x = -78.9382, label = "Duke", 
           color = "purple", size = 5) +
  theme_void() +
  theme(legend.position = "bottom") +
  labs(fill = "Death rate")
```

(3) What are limitations of your visualizations above?

Looking only at the plots, it may be difficult to differentiate counties. We are
also not adjusting for any of the COVID risk factors.

## Additional Resources

- [Simple features in `R`](https://r-spatial.github.io/sf/)
- [Coordinate references systems](https://www.nceas.ucsb.edu/sites/default/files/2020-04/OverviewCoordinateReferenceSystems.pdf)
- [Geographic data in `R`](https://geocompr.robinlovelace.net/spatial-class.html)
- [Leaflet for `R`](https://rstudio.github.io/leaflet/)
